Simulating cosmic variance in Weak Lensing: effects on parameter inferences

### 2015-11-05 Paper/paper.tex
finished first draft

\section{Introduction}
%
Weak Gravitational lensing is a promising cosmological probe for constraining the Dark Energy equation of state $w$, and has been considered by a range of past (CFHTLens), current (DES) and future (LSST) experiments. In an era where cosmology is becoming data driven, accurate numerical simulations of shear fields are becoming important for a number of reasons including the study of baryonic effects, non--Gaussian statistics, systematics... Building cosmological predictions from simulations naturally introduces fluctuations in the forecasts, mainly due to cosmic variance. These fluctuations have been shown to have non negligible effects on estimates of features covariance matrices and hence on parameter constraints (see \citep{DodelsonSchneider13}). This work studies this issue further, focusing on the number of independent $N$--body simulations that is necessary to run for obtaining accurate estimates of parameter constraints, with particular focus on $w$. This paper is organized as follows: in the first paragraph we briefly describe the shear simulation methods we used, as well as the formalism we adopted to compute cosmological parameter constraints. We then outline our main findings and discuss the results, as well as future prospects for continuing this study.

@@

\subsection{Shear field simulations}
\label{shearsim}
%
In this paragraph we describe how we constructed our shear field ensembles. Background galaxies at redshift $z_s$ are lensed by large scale structure between $z=0$ and $z_s$. The shape distortions due to the cosmic shear $\pmb{\gamma}$ can be computed in terms of the dark matter gravitational potential $\Phi(\bb{x},z)$. Because the evolution of $\Phi$ in redshift is non--linear, its evolution needs to be computed with numerical simulations. We make use of the public code Gadget2 \citep{Gadget2}, with which we run a sequence of dark--matter--only $N$--body simulations that track the evolution of the density fluctuations in the large scale structure of the universe. We assume a standard $\Lambda$CDM framework with $(\Omega_m,\Omega_\Lambda,h,w,\sigma_8,ns)=(0.26,0.74,0.72,-1,0.8,0.96)$ and we fix the comoving size of the simulation box to $240\mathrm{Mpc}/h$. We fill the box with $512^3$ particles, which correspond to a mass resolution of $10^{10}M_\odot$. We assume a uniform galaxy distribution at a constant redshift $z_s=2$ (at which the simulation box has an angular size of $\theta_{box}=3.5\mathrm{deg}$) and we disctretize the mass distribution between $z_s$ and the observer at $z=0$ with a sequence of 60 two dimensional lenses of thickness $80\mathrm{Mpc}/h$. The surface density on each lens is computed by projecting the three dimensional density measured from Gadget2 snapshots. We then apply the multi--lens--plane algorithm (see \citep{RayTracingHartlap,RayTracingJain} for example) to trace the deflections of $2048^2$ light rays arranged on a square grid of total size $\theta_{box}$, from $z=0$ to $z_s$. The light deflection calculations have been performed with our implementation of multi--lens--plane algorithm, which is part of the LensTools computing package \citep{LensTools} (released under the MIT license). Different realizations $r$ of the same shear field $\pmb{\gamma}_r(\pmb{\theta})$ can be generated with different choices of the lenses that lay between the observer and $z_s$. The randomization procedure we adopt is the following:

@@

\begin{equation}
\bbh{T} = (\bb{d}_0'^T\bbh{C}^{-1}\bb{d}_0')^{-1}\bb{d}_0'^T\bbh{C}^{-1}
\end{equation}
%
Where, for simplicity we assumed $\langle\bbh{d}\rangle=\bb{d}_0$, so that $\langle(\bbh{d}-\bb{d}_0)(\bbh{d}-\bb{d}_0)^T\rangle=\bb{C}$. Since the feature we consider is the shear-shear power spectrum, its derivatives $\bb{d}_0'$ with respect to the cosmological parameters can be evaluated by existing analytical codes. For this purpose we use the analytical code NICAEA (see \citep{Nicaea}). The noise in the covariance estimator (\ref{covest}) propagates all the way to the posterior (\ref{posteriorbayes}), the parameter estimate (\ref{estimatormean}) and its variance (\ref{estimatorcovariance}). \citep{DodelsonSchneider13} found that the variance $\h{\sigma}^2_\bb{p}=\mathrm{diag}(\bbh{\Sigma}_\bb{p})$ of each parameter $p$ that is estimated from (\ref{estimatorcovariance}) increases on average with $1/N_r$
\begin{equation}
\label{dodelsonscaling}
\langle\h{\sigma}_\bb{p}^2\rangle = \sigma_p^2\left(1+\frac{D}{N_r}\right)
\end{equation} 
%
where $\sigma^2_p$ is the true parameter variance and $D$ is what we call \textit{effective dimensionality} of the feature space. \citep{DodelsonSchneider13} found $D=N_b$ when every entry in the covariance matrix is subject to noise. In this work we argue that the feature ensemble from which the covariance estimator $\bbh{C}^{-1}$ is measured depends not only on the number of shear mock realizations $N_r$, but in principle also on the number of independent $N$--body simulations $N_s$, on which the randomization procedure we describe in \S~\ref{shearsim}. We propose a modification to the scaling relation (\ref{dodelsonscaling}) that takes the varying $N_s$ into account.

@@

@@

>>>>

\section{Introduction}
%
Weak Gravitational lensing is a promising cosmological probe for constraining the Dark Energy equation of state $w$, and has been considered by a range of past (CFHTLens \citep{cfht1,cfht2}), current (DES \citep{DES}) and future (LSST \citep{LSST}) experiments. In an era where cosmology is becoming data driven, accurate numerical simulations of shear fields are becoming important for a number of reasons including the study of baryonic effects \citep{BaryonXiuyuan}, non--Gaussian statistics \citep{PeaksJan,MinkJan,MinkPetri} and various systematics. Building cosmological predictions from simulations naturally introduces fluctuations in the forecasts, mainly due to cosmic variance. These fluctuations have been shown to have non negligible effects on estimates of features covariance matrices and hence on parameter constraints (see \citep{DodelsonSchneider13}). This work studies this issue further, focusing on the number of independent $N$--body simulations that is necessary to run for obtaining accurate estimates of parameter constraints, with particular focus on $w$. This paper is organized as follows: in the first paragraph we briefly describe the shear simulation methods we used, as well as the formalism we adopted to compute cosmological parameter constraints. We then outline our main findings and discuss the results, as well as future prospects for continuing this study.

@@

\subsection{Shear field simulations}
\label{shearsim}
%
In this paragraph we describe how we constructed our shear field ensembles. Background galaxies at redshift $z_s$ are lensed by large scale structure between $z=0$ and $z_s$. The shape distortions due to the cosmic shear $\pmb{\gamma}$ can be computed in terms of the dark matter gravitational potential $\Phi(\bb{x},z)$. Because the evolution of $\Phi$ in redshift is non--linear, its evolution needs to be computed with numerical simulations. We make use of the public code Gadget2 \citep{Gadget2}, with which we run a sequence of dark--matter--only $N$--body simulations that track the evolution of the density fluctuations in the large scale structure of the universe. We assume a standard $\Lambda$CDM framework with $(\Omega_m,\Omega_\Lambda,h,w,\sigma_8,n_s)=(0.26,0.74,0.72,-1,0.8,0.96)$ and we fix the comoving size of the simulation box to $240\mathrm{Mpc}/h$. We fill the box with $512^3$ particles, which correspond to a mass resolution of $10^{10}M_\odot$. We assume a uniform galaxy distribution at a constant redshift $z_s=2$ (at which the simulation box has an angular size of $\theta_{box}=3.5\mathrm{deg}$) and we disctretize the mass distribution between $z_s$ and the observer at $z=0$ with a sequence of 60 two dimensional lenses of thickness $80\mathrm{Mpc}/h$. The surface density on each lens is computed by projecting the three dimensional density measured from Gadget2 snapshots. We then apply the multi--lens--plane algorithm (see \citep{RayTracingHartlap,RayTracingJain} for example) to trace the deflections of $2048^2$ light rays arranged on a square grid of total size $\theta_{box}$, from $z=0$ to $z_s$. The light deflection calculations have been performed with our implementation of multi--lens--plane algorithm, which is part of the LensTools computing package \citep{LensTools} (released under the MIT license). Different realizations $r$ of the same shear field $\pmb{\gamma}_r(\pmb{\theta})$ can be generated with different choices of the lenses that lay between the observer and $z_s$. The randomization procedure we adopt is the following:

@@

\begin{equation}
\bbh{T} = (\bb{d}_0'^T\bbh{C}^{-1}\bb{d}_0')^{-1}\bb{d}_0'^T\bbh{C}^{-1}
\end{equation}
%
Where, for simplicity we assumed $\langle\bbh{d}\rangle=\bb{d}_0$, so that $\langle(\bbh{d}-\bb{d}_0)(\bbh{d}-\bb{d}_0)^T\rangle=\bb{C}$. Since the feature we consider is the shear-shear power spectrum, its derivatives $\bb{d}_0'$ with respect to the cosmological parameters can be evaluated by existing analytical codes. For this purpose we use the analytical code NICAEA (see \citep{Nicaea}). The noise in the covariance estimator (\ref{covest}) propagates all the way to the posterior (\ref{posteriorbayes}), the parameter estimate (\ref{estimatormean}) and its variance (\ref{estimatorcovariance}), through the transformation matrix $\bbh{T}$. \citep{DodelsonSchneider13} found that the variance $\h{\sigma}^2_\bb{p}=\mathrm{diag}(\bbh{\Sigma}_\bb{p})$ of each parameter $p$ that is estimated from (\ref{estimatorcovariance}) increases on average with $1/N_r$
\begin{equation}
\label{dodelsonscaling}
\langle\h{\sigma}_\bb{p}^2\rangle = \sigma_p^2\left(1+\frac{D}{N_r}\right)
\end{equation} 
%
where $\sigma^2_p$ is the true parameter variance and $D$ is what we call \textit{effective dimensionality} of the feature space. \citep{DodelsonSchneider13} found $D=N_b$ when every entry in the covariance matrix is subject to noise. In this work we argue that the feature ensemble from which the covariance estimator $\bbh{C}^{-1}$ is measured depends not only on the number of shear mock realizations $N_r$, but in principle also on the number of independent $N$--body simulations $N_s$, on which the randomization procedure we describe in \S~\ref{shearsim}. We propose a modification to the scaling relation (\ref{dodelsonscaling}) that takes the varying $N_s$ into account.

@@

In this section we discuss our main findings. Figure \ref{ps_pdf} shows that, although different choices of $N_s$ do not seem to affect the power spectrum PDF at large scales (top two panels), there are some qualitative differences on small scales (bottom two panels), on which a shear ensemble built with a small $N_s$ might not have the same statistical behavior of ensembles built with large $N_s$. Figure \ref{ps_var} shows the variance of convergence power spectrum computed from different ensembles, in units of the Gaussian expectation. We find that, even with $N_s=1$, our results are in good agreement with the ones obtained by \citep{Sato12}, which used $N_s=400$. This fact by itself is not sufficient to conclude that $N_s$ does not have an effect on parameter inferences, since these depend on the cross band covariances.

In Figure \ref{wvar_nr} we examine how the $w$ constraint depends on the number of mocks used to estimate the covariance, and we find good agreement with \citep{DodelsonSchneider13}. We determine the effective dimensionality of the feature space, $D$, with a linear regression using equation (\ref{ourscaling}) and we find that this depends strongly on how the noise is distributed across the covariance matrix estimator $\bbh{C}$. In particular we find that when every element in the covariance is subject to noise $D$ is close to the number of bands used $N_b=15$, but when we artificially zero out the off diagonal elements of $\bbh{C}$, the effective dimensionality drops to $D\approx2.5$. Finally, Figure \ref{wvar_ns} shows the dependence of the dark energy constraint with $N_s$ and $D$. We find that, in the range $N_s\in[1,200]$ the $w$ inferred variance has sub--percent fluctuations for both the diagonal and full $\bbh{C}$ cases, while there is a $\sim$10\% change in $\sigma^2_0(N_s,w)$ when the effective dimensionality $D$ is different. Finally, the different lines in Figure \ref{wvar_ns} show that equation (\ref{ourscaling}) is a good model for the variance scaling.

@@

In this work we examine the effect of $N$--body simulations based shear ensembles on forecasted cosmological constraints. Our main results can be summarized as follows:

\begin{itemize}
\item The variance scaling relation proposed by \citep{DodelsonSchneider13} is a good approximation regardless of $N_s$
\item The fluctuations in the $w$ forecasted variance due to varying $N_s$ are sub--percent, and this effect is sub--dominant with respect of varying the effective dimensionality of the feature space, which in the end is controlled by how the noise is distributed in the covariance estimator
\end{itemize}
%
Future prospects of this work involve extending this analysis to more general feature spaces, such as the ones that characterize non--Gaussian statistics such as shear peaks and higher moments of $\kappa$ fields. In order to scale these considerations to future surveys such as LSST, it is necessary to see if our findinds hold when challenged by larger and higher resolutions $N$--body simulations.
